---
title: "Homework Assignment 1"
author: "Ian Cooper"
date: 'Due EoD Wednesday Nov 3, 2021'
output:
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
  pdf_document: default
    
---

## HW-1 Instructions

> Our class emphasizes clear communication of data analysis results to non-technical audiences. I expect your HTML output documents to be readable and well formatted. I expect you to put ample comments in your R code to make the code understandable. Along with accuracy of results, I am looking for quality of presentation as well. To complete this assignment, follow these steps:

1.  Download the `HW1.Rmd` file from Canvas.

2.  Open `HW1.Rmd` in RStudio. Replace the "Your Name Here" text in the `author:` field with your name.

3.  Supply your solutions to the homework by editing `HW1.Rmd`.

4.  Run your code in the Console and Knit HTML frequently to check for errors.

5.  You may find it easier to solve a problem by interacting only with the Console at first.

6.  When you have completed the homework please check that your knits correctly when you click `Knit HTML`.

7.  "Push" all your local changes to the GitHub Repo for the project.

8.  Submit your RMD file, the HTML output and a link to your GitHub Repo on Canvas.

## Data frame basics

We will work with the nycflights dataset. **Please be sure to keep the data file in the same directory as the RMD file - your project directory.** and make that directory the working directory.

First - make sure that you are reading the data correctly and doing all the data cleaning steps that we did in class. Feel free to go beyond our class work in cleaning the data if you feel the need.

```{r}
#Read the data file 
nyc <- read.csv("nycflights.csv")
nyc$X <- NULL
head(nyc)

# Structure of the data
str(nyc)

# Summary of the data
summary(nyc)


```

### Data Cleaning

We will clean our data now

```{r}
#Lets convert things to factors
nyc$flight <- as.factor(nyc$flight)

#Now convert the following to factors 
#Origin, dest, carrier, tailnum
nyc$origin <- as.factor(nyc$origin)
nyc$dest <- as.factor(nyc$dest)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$carrier <- as.factor(nyc$carrier)

nyc$flightdate <- as.Date(nyc$time_hour,"%Y-%m-%d %H:%M:%S")
str(nyc)

summary(nyc)

```

I now would like you to answer the following - all in their own separate R code blocks.

### Data Exploration

Let's first do some simple exploration of this data.

-   How many airlines are there? (Hint: `levels` and `length` can be useful here)
```{r}
nlevels(nyc$carrier)
```

-   How many flights there were by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)
```{r}
nycoo <- nyc[ nyc$carrier == "OO", ]
nrow(nycoo)
```

-   How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)
```{r}
# I am making the assumption that "shortest flight" means flight duration 
min(nyc$air_time, na.rm = TRUE)

```

The shortest flight you can take out of any NYC airport is `r min(nyc$air_time, na.rm = TRUE)` mins. This is the shortest time and not the shortest distance.

-   How many flights there were by United Airlines (code: UA) on Jan 12th 2013?
```{r}
nycUAJan12 <- nyc[ nyc$carrier == "UA" &  nyc$month == 1 & nyc$day == 12, ]
nrow(nycUAJan12)
```

### Arrival Delay

Lets focus on Arrival Delay.

-   What was the average arrival delay for all airports and all airlines combined in 2013?
```{r}
round(mean(nyc$arr_delay, na.rm = TRUE), digits = 2)
```

-   Whats was the median arrival delay for all airports and all airlines combined in 2013?
```{r}
median(nyc$arr_delay, na.rm = TRUE)

```

Based on your answers to the two questions above, what can you say about the distribution of arrival delays? Provide your answer in a text paragraph form.

Because the median arrival delay is -5 while the average is 6.9, we can say that there is likely a long right-hand tail of arrival delay observations that is pushing the average well above the median, while the left-hand tail is more condensed. In essence, while most flights are on time or even early, there are a few flights that are significantly delayed, pushing the average up.

### Airline Performance

Lets see if all airlines are equally terrible as far as flight arrival delays are concerned. For this question you will have to make sure that airline column is coded as a factor.

-   Calculate average arrival delays by airline (Hint: look up the command `tapply`)
```{r}
arr_delay <- nyc$arr_delay
carrier <- nyc$carrier
delaybyairline <- sort(round(tapply(arr_delay, carrier, mean, na.rm = TRUE), digits = 1), decreasing = TRUE)
delaybyairline
```

-   Draw a Bar Plot of Average Arrival Delays for all the Airlines (Hint: command for making a Bar Plot is simply `barplot`)
```{r}
barplot(delaybyairline, las = 2)
```

-   Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay? Provide answer to this question in a text paragraph form using **inline R code**.

  The airline with the highest average arrival delay is `r delaybyairline[1]` while the airline with the smallest is `r delaybyairline[16]`. `r names(delaybyairline[15])` and `r names(delaybyairline[16])` both have negative average delays. 

### Air Gain

Create a new column named airgain such that airgain = (departure delay - arrival delay) : this is the amount of delay a flight made up while in air.
```{r}
nyc$airgain <- nyc$dep_delay - nyc$arr_delay
```

a)  Explore airgain data - calculate suitable descriptive statistics and appropriate graphics to better understand this data. This part is open ended - you do what you feel works best for you.
```{r}
#Summary of the data
summary(nyc$airgain)

#Histogram of the data
hist(nyc$airgain[nyc$airgain > -50 & nyc$airgain < 50])

#Finding the mean airgain
round(mean(nyc$airgain, na.rm = TRUE), digits = 2)

#Sorting the airlines by average airgain in decreasing order
airgainbyairline <- sort(round(tapply(nyc$airgain, carrier, mean, na.rm = TRUE), digits = 1), decreasing = TRUE)
airgainbyairline

#Making a barplot of the average airgains by airline in decreasing order
barplot(airgainbyairline, las = 2)

#Finding the average airgain by airport in decreasing order
airgainbyairport <- sort(round(tapply(nyc$airgain, nyc$origin, mean, na.rm = TRUE), digits = 1), decreasing = TRUE)
airgainbyairport

#Making a barplot of the average airgain by airport in decreasing order
barplot(airgainbyairport)
```

b)  Answer the questions:

-   do airlines actually gain any time when in air on average?
```{r}
round(mean(nyc$airgain, na.rm = TRUE), digits = 2)
```

Yes, on average, airlines flying out of NYC gain `r round(mean(nyc$airgain, na.rm = TRUE), digits = 2)` mins.

-   Calculate average airgain for different airlines - which airlines do a better job, which do a worse job?
```{r}
barplot(airgainbyairline, las = 2)
```

As you can see in the above barplot, some airlines perform better than others in terms of their average airgains. While the average airgain for all airlines is `r round(mean(nyc$airgain, na.rm = TRUE), digits = 1)` mins, the best performing airline is `r names(airgainbyairline[1])` with an average of `r max(airgainbyairline)` mins, while the worst performing is `r names(airgainbyairline[16])` with an average of `r min(airgainbyairline)` mins.

### Merging Data Frames

You can get detailed information about the physical planes in our dataset in this file: `planes.csv`. Download and save this file in your project directory.

a)  Read the `planes.csv` file using `read.csv` command. Do any data cleaning necessary.
```{r}
#Read the file
planes <- read.csv("planes.csv")
str(planes)
summary(planes)


#Factorize the following: Manufacturer, Type, Model, Engine and Tailnumber
planes$manufacturer <- as.factor(planes$manufacturer)
planes$type <- as.factor(planes$type)
planes$tailnum <- as.factor(planes$tailnum)

```

b)  Merge the flights data and the planes data using the `merge` command. You should do the merge on the common column named `tailnum`. *getting this right may need some trial and error and getting some help*.
```{r}
#Merging the files
nycjoined <- merge(nyc, planes, by = "tailnum")
nycjoined$X <- NULL
str(nycjoined)

```

c)  Now that you have a merged dataset, think of what interesting questions that you can ask that can be answered using the merged dataset. You are asked to pose five interesting questions and answer them. (For example: who are the top 10 manufacturers of planes that fly out of NYC airports?) **Be creative. Be bold. Ask questions that you would want to know answers to even if you were not doing this for a HW.**


- Make a histogram of planes with more than 50 seats flying out of NYC. 
```{r}
hist(nycjoined$seats[nycjoined$seats > 50])
```

- What is the average number of seats for a plane flying out of NYC?
```{r}
round(mean(nycjoined$seats, na.rm = TRUE), digits = 0)
```

- How old is the oldest plane used in 2013?

The oldest plane used in 2013 was `r 2013 - min(nycjoined$year.y, na.rm = TRUE) ` years old.

- Filter the average size of planes (number of seats) by airport.
```{r}
SeatsByAirport <- sort(round(tapply(nycjoined$seats, nycjoined$origin, mean, na.rm = TRUE), digits = 0), decreasing = TRUE)
SeatsByAirport
```
- Filter average airgain by manufacturer. What manufacturer has the highest average airgain?
```{r}
AirgainByManufacturer <- sort(round(tapply(nycjoined$airgain, nycjoined$manufacturer, mean, na.rm = TRUE), digits = 1), decreasing = TRUE)
AirgainByManufacturer
```
`r names(AirgainByManufacturer[1])` has the highest average airgain.


### Making Your HTML Look Nice

We want our report to be good looking, professional documents. To that end, I am asking you to do the following:

-   Have a floating table of contents

-   Include code folding in your output. You can find more about code folding here: <https://bookdown.org/yihui/rmarkdown/html-document.html#code-folding>

That's it. Once you are done, make sure everything works and knits well and then you can push your changes to the GitHub repo and upload the RMD file and the html output to Canvas.

**Have Fun!**

Sanjeev
